Laplacian envelope

### 2021-07-31 Notes/laplace.tex
rearranged figs

First, let's look at the variances, remembering that the states and plasticity events of each synapse are independent and identically distributed:
%
\begin{equation}\label{eq:noise}
\begin{aligned}
  \var(\synid\cdt\syn(t))
    &= \sum_{\alpha\beta} \av{\synid^\alpha \syn^\alpha(t) \synid^\beta \syn^\beta(t)}
    - \Big(\sum_{\alpha} \av{\synid^\alpha \syn^\alpha(t)} \Big)^2 \\
    &= \sum_{\alpha} \av{(\synid^\alpha)^2(\syn^\alpha(t))^2}
    + \sum_{\alpha\neq\beta} \av{\synid^\alpha\syn^\alpha(t)}\av{\smash{\synid^\beta}\smash{\syn^\beta}(t)}
    %\\&\phantom{= \sum_{\alpha}  \av{(\synid^\alpha)^2(\syn^\alpha(t))^2}}
    - \Big(\sum_{\alpha} \av{\synid^\alpha \syn^\alpha(t)} \Big)^2  \\
    &= N\av{1}
    + N(N-1)\av{\synid^1 \syn^1(t)}^2
    - N^2\av{\synid^1 \syn^1(t)}^2 \\
    &= N(1-\av{\synid^1 \syn^1(t)}^2),
\end{aligned}
\end{equation}
%
where we used \(\syn^\alpha=\pm1\).
We can compute \(\var(\synid\cdt\syn(\infty))\) by taking \(t\to\infty\).

@@

The derivatives, for \(m \neq n\) are
%
\begin{equation}\label{eq:finitederivs}
\begin{aligned}
  \frac{1}{f^\nu} \pdiff{\CL}{\MMdm^\nu_{mn}} = & \,
    \eqm_m \prn{ \brk{\thbm_m(s) - \thbm_n(s)}
     + \prn{\mathbf{c}_m(s) + (-1)^\nu} \brk{\etwm_m(s) - \etwm_n(s)} } \\
     &+ \mu^\nu_{mn} -  \mu^\nu_{mm}
     + \lambda \eqm_m (\etwm_{m}(0)-\etwm_{n}(0)) 
    = 0.
\end{aligned}
\end{equation}
%
We can solve for the KKT multipliers by noting that
%
\begin{equation*}
  \mu^\nu_{ij} \MMdm^\nu_{ij} = 0,
  \qquad \text{and} \qquad
  \sum_{j: j \neq i} \mu^\nu_{ii} \MMdm^\nu_{ij} 
      = \mu^\nu_{ii} (1 - \MMdm^\nu_{ii}) = \mu^\nu_{ii}.
\end{equation*}
%
Then, with some new vectors defined:
%
\begin{equation}\label{eq:kktsol}
\begin{aligned}
  \mu^\nu_{mm} &= - \eqm_m \prn{ \albm^\nu_m(s) 
    + \lambda \ombm^\nu_m(0) 
    + \prn{{c}_m(s) + (-1)^\nu} \ombm^\nu_m(s)
    }, \\
  \mu^\nu_{mn} &= \eqm_m \big( 
      \brk{ \thbm_n(s) - \thbm_m(s) - \albm^\nu_m(s) }
    + \lambda \brk{ \etwm_n(0) - \etwm_m(0) - \ombm^\nu_m(0) } 
  \\ &\phantom{= \eqm_m \big( }
    + \brk{{c}_m(s) + (-1)^\nu} 
        \brk{ \etwm_n(s) - \etwm_m(s) - \ombm^\nu_m(s) }
     \big), \\
  &\text{where} \quad 
  \alb^\nu(s) = (\MMd^\nu - \I) \thb(s), \qquad
  \omb^\nu(s) = (\MMd^\nu - \I) \etw(s).
\end{aligned}
\end{equation}
%
If we note that we can write \( \MMd^\nu - \I = \prn{\frg + (-1)^\nu \enc} / 2f^\nu \), we find that
%
\begin{equation}\label{eq:diagKKT}
\begin{gathered}
  \alb^\nu(s) = \frac{\delta \mathbf{a}(s) + (-1)^\nu \enc \thb(s)}
                     {2f^\nu},
  \qquad 
  \omb^\nu(s) = \frac{\delta \w + s \delta \etw(s) - (-1)^\nu \mathbf{a}(s)}
                     {2 f^\nu}, \\
\begin{aligned}
  \mu^\nu_{mm} &= \frac{\eqm_m}{2f^\nu} 
    \left[ 
      \hat{A}(s) - c_m(s) (\delta \wm_m + s \delta \etwm_m(s))
      - \lambda \delta w_m
    \right. \\ &\phantom{= \frac{\eqm_m}{2f^\nu} }
    \left.
      + (-1)^\nu \prn{ 
        c_m(s) a_m(s) - [\encm \thbm(s)]_m
        - \delta \wm_m - s \delta \etwm_m(s)
        + \lambda a_m(0)
      }
    \right], \\
\end{aligned}
\end{gathered}
\end{equation}
%
where \(\delta\) indicates mean subtraction and \(\eq \mathbf{a}(s) = \hat{A}(s)\).

@@

However, we do have numerical evidence that the model that maximises the Laplace transform does have the serial topology.
In \cref{fig:envnum} we see the result of numerical maximisation of \cref{eq:finitelagrangian}.
We see that restricting to the serial topology does not make the performance any worse.
In fact, it slightly improves the performance, but this can be ascribed to the numerical optimisation problem being easier.
Assuming that the optimiser did find the global maximum, this provides evidence for our conjecture that the model that maximises the Laplace transform must have the serial topology.

Note that models with the serial topology must satisfy detailed balance.
Therefore, this conjecture implies that the square-root bound of \cref{sec:sqrt} applies, implying the stricter envelope of \cref{eq:envDB}.
In \cref{fig:envnum}, we can see that this envelope is close to tight, being off by a constant factor.

\begin{figure}[tb]
  \centering
  % Requires \usepackage{graphicx}
  \includegraphics[width=0.8\linewidth]{LenvNum.pdf}\\
  \caption[Numerical envelope for the signal-to-noise ratio]
  {Numerical envelope for the signal-to-noise ratio, \eqref{eq:snrb}. 
  \Cref{eq:finitelagrangian,eq:KTfinite,eq:finitederivs} are solved numericallly with \(f\potdep=\frac{1}{2}\), the number of synapses \(N=1\) and the number of states \(M=10\).
  The orange line indicates maximisation of the memory curve over the space of all matrices satisfying the constraints \eqref{eq:constr}.
  The green line indicates maximisation over models with the serial topology, \ie potentiation only moves up one state and depression only moves down one state.
  The dotted blue line indicates the proven envelope, \cref{eq:env} and the solid blue line is the conjectured envelope assuming detailed balance, see \cref{eq:envDB}.}\label{fig:envnum}
\end{figure}

@@

The derivatives, for \(m \neq n\) are
%
\begin{equation}\label{eq:lshiftderivs}
\begin{aligned}
  \frac{1}{f^\nu} \pdiff{\CL}{\MMdm^\nu_{mn}} = & \,
    \eqm_m \prn{ \brk{\thbm_m - \thbm_n}
     + \prn{{c}_m + (-1)^\nu + \lambda} \brk{\etwm_m - \etwm_n} } \\
     &+ \mu^\nu_{mn} -  \mu^\nu_{mm}
     - \sum_{\mu ij} \mu^\mu_{ij} s f^\mu \eqm_m (\fptbm_{mj} - \fptbm_{nj}) \eqm_j
    = 0.
\end{aligned}
\end{equation}
%
We can absorb the new terms by defining
%
\begin{equation*}
\begin{aligned}
  \shift{\thb}^\nu(s) = \thb(0) - s f^\nu \fptb(0) \Eq (\boldsymbol{\mu}^\nu)\trans \onev,
  \qquad
  \shift{\alb}^\nu(s) &= (\MMd^\nu + (s - 1) \I - s \onev \eq) \shift{\thb}^\nu(s),
  \\
  \shift{\omb}^\nu(s) &= (\MMd^\nu + (s - 1) \I - s \onev \eq) \etw(0).
\end{aligned}
\end{equation*}
%
We note that
%
\begin{equation*}
  \mu^\nu_{ij} (\MMdm^\nu_{ij} - s \eqm_j) = 0,
  \qquad \text{and} \qquad
  \sum_{j: j \neq i} \mu^\nu_{ii} (\MMdm^\nu_{ij} - s \eqm_j)
      = \mu^\nu_{ii} (1 - \MMdm^\nu_{ii} - s (1 - \eqm_i)) = \mu^\nu_{ii},
\end{equation*}
%
then we can ``solve'' for the KKT multipliers:
%
\begin{equation}\label{eq:lshiftsol}
\begin{aligned}
  \mu^\nu_{mm} &= - \eqm_m \prn{ \shift{\albm}^\nu_m(s) 
    + \brk{ {c}_m + (-1)^\nu + \lambda } \shift{\ombm}^\nu_m(s)
    }, \\
  \mu^\nu_{mn} &= \eqm_m \prn{ 
      \brk{ \shift{\thbm}^\nu_n(s) - \shift{\thbm}^\nu_m(s) - \shift{\albm}^\nu_m(s) }
  % \\ & \phantom{= - \eqm_m \big( } 
    + \brk{ {c}_m + (-1)^\nu + \lambda } 
        \brk{ \etwm_n - \etwm_m - \shift{\ombm}^\nu_m(s) }
     }, \\
\end{aligned}
\end{equation}
%
although there are KKT multipliers hiding in \(\shift{\thb}^\nu(s)\) and \(\shift{\alb}^\nu(s)\).

@@

Let us define the following
%
\begin{equation}\label{eq:degenerate}
  \dgn_{m} = \thb + c_m \etw,
  \quad
  \dgn^\nu_m = \dgn_m + (-1)^\nu \etw,
  \quad \implies \;
  \frac{1}{f^\nu} \pdiff{\CL}{\MMdm^\nu_{mn}} = \eqm_m (\dgnm^\nu_{m,m} - \dgnm^\nu_{m,n}) + \mu^\nu_{mn} - \mu^\nu_{mm}.
\end{equation}
%
They should be thought of as collections of vectors rather than matrices.
We can also write \cref{eq:diagKKT} as
%
\begin{equation}\label{eq:diagKKTz}
\begin{aligned}
  \mu^\nu_{mm} &= \frac{\eqm_m}{2f^\nu} 
    \left[ 
      \hat{A}(s) - c_m(s) (\delta \wm_m + s \delta \etwm_m(s))
      - \lambda \delta w_m
    \right. \\ &\phantom{= \frac{\eqm_m}{2f^\nu} }
    \left.
      - (-1)^\nu \prn{ 
        [\encm \dgnm(s)]_m
        + \delta \wm_m + s \delta \etwm_m(s)
        - \lambda a_m(0)
      }
    \right], \\
\end{aligned}
\end{equation}
%

@@

\begin{figure}[tb]
  \centering
  % Requires \usepackage{graphicx}
  \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.4\linewidth]{LenvProven.pdf}}\label{fig:provengen}
  \item\aligntop{\includegraphics[width=0.4\linewidth]{LenvConj.pdf}}\label{fig:provendetbal}
  \end{myenuma}
  \caption[Proven envelope for the signal-to-noise ratio]
  {(\ref{fig:provengen}) The memory curve envelope that we can prove, \cref{eq:env}.
  (\ref{fig:provendetbal}) Conjectured memory curve envelope when we assume the optimal models satisfy detailed balance, \cref{eq:envDB}.
  The number of synapses \(N=1\) and the number of states \(M=10\).
  The dashed lines indicate the initial SNR bound, \eqref{eq:initbnd}, and the area bound, \eqref{eq:areabnd}, from \cite{Lahiri2013synapse}.}\label{fig:envproven}
\end{figure}

@@

>>>>

First, let's look at the variances, remembering that the states and plasticity events of each synapse are independent and identically distributed:
%
\begin{equation}\label{eq:noise}
\begin{aligned}
  \var(\synid\cdt\syn(t))
    &= \sum_{\alpha\beta} \av{\synid^\alpha \syn^\alpha(t) \synid^\beta \syn^\beta(t)}
    - \Big(\sum_{\alpha} \av{\synid^\alpha \syn^\alpha(t)} \Big)^2 \\
    &= \sum_{\alpha} \av{(\synid^\alpha)^2(\syn^\alpha(t))^2}
    + \sum_{\alpha\neq\beta} \av{\synid^\alpha\syn^\alpha(t)}\av{\smash{\synid^\beta}\smash{\syn^\beta}(t)}
    %\\&\hphantom{= \sum_{\alpha}  \av{(\synid^\alpha)^2(\syn^\alpha(t))^2}}
    - \Big(\sum_{\alpha} \av{\synid^\alpha \syn^\alpha(t)} \Big)^2  \\
    &= N\av{1}
    + N(N-1)\av{\synid^1 \syn^1(t)}^2
    - N^2\av{\synid^1 \syn^1(t)}^2 \\
    &= N(1-\av{\synid^1 \syn^1(t)}^2),
\end{aligned}
\end{equation}
%
where we used \(\syn^\alpha=\pm1\).
We can compute \(\var(\synid\cdt\syn(\infty))\) by taking \(t\to\infty\).

@@

The derivatives, for \(m \neq n\) are
%
\begin{equation}\label{eq:finitederivs}
\begin{aligned}
  \frac{1}{f^\nu} \pdiff{\CL}{\MMdm^\nu_{mn}} = & \,
    \eqm_m \prn{ \brk{\thbm_m(s) - \thbm_n(s)}
     + \prn{\mathbf{c}_m(s) + (-1)^\nu} \brk{\etwm_m(s) - \etwm_n(s)} } \\
     &+ \mu^\nu_{mn} -  \mu^\nu_{mm}
     + \lambda \eqm_m (\etwm_{m}(0)-\etwm_{n}(0)) 
    = 0.
\end{aligned}
\end{equation}
%
We can solve for the KKT multipliers by noting that
%
\begin{equation*}
  \mu^\nu_{ij} \MMdm^\nu_{ij} = 0,
  \qquad \text{and} \qquad
  \sum_{j: j \neq i} \mu^\nu_{ii} \MMdm^\nu_{ij} 
      = \mu^\nu_{ii} (1 - \MMdm^\nu_{ii}) = \mu^\nu_{ii}.
\end{equation*}
%
Then, with some new vectors defined:
%
\begin{equation}\label{eq:kktsol}
\begin{aligned}
  \mu^\nu_{mm} &= - \eqm_m \prn{ \albm^\nu_m(s) 
    + \lambda \ombm^\nu_m(0) 
    + \prn{{c}_m(s) + (-1)^\nu} \ombm^\nu_m(s)
    }, \\
  \mu^\nu_{mn} &= \eqm_m \big( 
      \brk{ \thbm_n(s) - \thbm_m(s) - \albm^\nu_m(s) }
    + \lambda \brk{ \etwm_n(0) - \etwm_m(0) - \ombm^\nu_m(0) } 
  \\ &\hphantom{= \eqm_m \big( }
    + \brk{{c}_m(s) + (-1)^\nu} 
        \brk{ \etwm_n(s) - \etwm_m(s) - \ombm^\nu_m(s) }
     \big), \\
  \text{where} \quad 
  \alb^\nu(s) &= (\MMd^\nu - \I) \thb(s), \qquad
  \omb^\nu(s) = (\MMd^\nu - \I) \etw(s).
\end{aligned}
\end{equation}
%
If we note that we can write \( \MMd^\nu - \I = \prn{\frg + (-1)^\nu \enc} / 2f^\nu \), we find that
%
\begin{equation}\label{eq:diagKKT}
\begin{aligned}
  \alb^\nu(s) &= \frac{\delta\mathbf{a}(s) + (-1)^\nu \enc \thb(s)}
                     {2f^\nu},
  \qquad 
  \omb^\nu(s) = \frac{\delta\w + s \delta\etw(s) - (-1)^\nu \mathbf{a}(s)}
                     {2 f^\nu}, 
  \\
  \mu^\nu_{mm} &= \frac{\eqm_m}{2f^\nu} 
    \left[ 
      \hat{A}(s) - c_m(s) (\delta\wm_m + s \delta\etwm_m(s))
      - \lambda \delta w_m
    \right. \\ &\hphantom{= \frac{\eqm_m}{2f^\nu} \big[}
    \left. {}
      + (-1)^\nu \prn{ 
        c_m(s) a_m(s) - [\encm \thbm(s)]_m
        - \delta\wm_m - s \delta\etwm_m(s)
        + \lambda a_m(0)
      }
    \right], 
\end{aligned}
\end{equation}
%
where \(\delta\) indicates mean subtraction and \(\eq \mathbf{a}(s) = \hat{A}(s)\).

@@

However, we do have numerical evidence that the model that maximises the Laplace transform does have the serial topology.
In \cref{fig:envnum} we see the result of numerical maximisation of \cref{eq:finitelagrangian}.
We see that restricting to the serial topology does not make the performance any worse.
In fact, it slightly improves the performance, but this can be ascribed to the numerical optimisation problem being easier.
Assuming that the optimiser did find the global maximum, this provides evidence for our conjecture that the model that maximises the Laplace transform must have the serial topology.
The models found numerically are described in \cref{fig:envvid}.

\begin{figure}[tb]
  \centering
  % Requires \usepackage{graphicx}
  \includegraphics[width=0.8\linewidth]{LenvNumOnly.pdf}\\
  \caption[Numerical envelope for the signal-to-noise ratio]
  {Numerical envelope for the signal-to-noise ratio, \eqref{eq:snrb}. 
  \Cref{eq:finitelagrangian,eq:KTfinite,eq:finitederivs} are solved numerically with \(f\potdep=\frac{1}{2}\), the number of synapses \(N=1\) and the number of states \(M=10\).
  The orange line indicates maximisation of the memory curve over the space of all matrices satisfying the constraints \eqref{eq:constr}.
  The green line indicates maximisation over models with the serial topology, \ie potentiation only moves up one state and depression only moves down one state.
  }\label{fig:envnum}
\end{figure}

@@

The derivatives, for \(m \neq n\) are
%
\begin{equation}\label{eq:lshiftderivs}
\begin{aligned}
  \frac{1}{f^\nu} \pdiff{\CL}{\MMdm^\nu_{mn}} = & \,
    \eqm_m \prn{ \brk{\thbm_m - \thbm_n}
     + \prn{{c}_m + (-1)^\nu + \lambda} \brk{\etwm_m - \etwm_n} } \\
     &+ \mu^\nu_{mn} -  \mu^\nu_{mm}
     - \sum_{\mu ij} \mu^\mu_{ij} s f^\mu \eqm_m (\fptbm_{mj} - \fptbm_{nj}) \eqm_j
    = 0.
\end{aligned}
\end{equation}
%
We can absorb the new terms by defining
%
\begin{equation*}
\begin{aligned}
  \shift{\thb}^\nu(s) = \thb(0) - s f^\nu \fptb(0) \Eq (\boldsymbol{\mu}^\nu)\trans \onev,
  \qquad
  \shift{\alb}^\nu(s) &= (\MMd^\nu + (s - 1) \I - s \onev \eq) \shift{\thb}^\nu(s),
  \\
  \shift{\omb}^\nu(s) &= (\MMd^\nu + (s - 1) \I - s \onev \eq) \etw(0).
\end{aligned}
\end{equation*}
%
We note that
%
\begin{equation*}
  \mu^\nu_{ij} (\MMdm^\nu_{ij} - s \eqm_j) = 0,
  \qquad \text{and} \qquad
  \sum_{j: j \neq i} \mu^\nu_{ii} (\MMdm^\nu_{ij} - s \eqm_j)
      = \mu^\nu_{ii} (1 - \MMdm^\nu_{ii} - s (1 - \eqm_i)) = \mu^\nu_{ii},
\end{equation*}
%
then we can ``solve'' for the KKT multipliers:
%
\begin{equation}\label{eq:lshiftsol}
\begin{aligned}
  \mu^\nu_{mm} &= - \eqm_m \prn{ \shift{\albm}^\nu_m(s) 
    + \brk{ {c}_m + (-1)^\nu + \lambda } \shift{\ombm}^\nu_m(s)
    }, \\
  \mu^\nu_{mn} &= \eqm_m \prn{ 
      \brk{ \shift{\thbm}^\nu_n(s) - \shift{\thbm}^\nu_m(s) - \shift{\albm}^\nu_m(s) }
  % \\ & \hphantom{= - \eqm_m \big( } 
    + \brk{ {c}_m + (-1)^\nu + \lambda } 
        \brk{ \etwm_n - \etwm_m - \shift{\ombm}^\nu_m(s) }
     }, \\
\end{aligned}
\end{equation}
%
although there are KKT multipliers hiding in \(\shift{\thb}^\nu(s)\) and \(\shift{\alb}^\nu(s)\).

@@

Let us define the following
%
\begin{equation}\label{eq:degenerate}
  \dgn_{m} = \thb + c_m \etw,
  \quad
  \dgn^\nu_m = \dgn_m + (-1)^\nu \etw,
  \quad \implies \;
  \frac{1}{f^\nu} \pdiff{\CL}{\MMdm^\nu_{mn}} = \eqm_m (\dgnm^\nu_{m,m} - \dgnm^\nu_{m,n}) + \mu^\nu_{mn} - \mu^\nu_{mm}.
\end{equation}
%
They should be thought of as collections of vectors rather than matrices.
We can also write \cref{eq:diagKKT} as
%
\begin{equation}\label{eq:diagKKTz}
\begin{aligned}
  \mu^\nu_{mm} &= \frac{\eqm_m}{2f^\nu} 
    \left[ 
      \hat{A}
      - (c_m + (-1)^\nu) (\delta \wm_m + s \delta \etwm_m)
      - \lambda \delta w_m
    % \right. \\ &\hphantom{= \frac{\eqm_m}{2f^\nu} }
    % \left.
      - (-1)^\nu \prn{ 
        [\encm \dgnm_m]_m
        - \lambda a_m(0)
      }
    \right],
\end{aligned}
\end{equation}
%
with all quantities evaluated at \(s\) unless labelled otherwise.

@@

\begin{figure}[tb]
  \centering
  % Requires \usepackage{graphicx}
  \begin{myenuma}
  \item\aligntop{\includegraphics[width=0.4\linewidth]{LenvProven.pdf}}\label{fig:provengen}
  \item\aligntop{\includegraphics[width=0.4\linewidth]{LenvConj.pdf}}\label{fig:provendetbal}
  \end{myenuma}
  \caption[Proven and conjectured envelopes for the signal-to-noise ratio]
  {Proven and conjectured envelopes for the signal-to-noise ratio.
  (\ref{fig:provengen}) The memory curve envelope that we can prove, \cref{eq:env}.
  (\ref{fig:provendetbal}) Conjectured memory curve envelope when we assume the optimal models satisfy detailed balance, \cref{eq:envDB}.
  The number of synapses \(N=1\) and the number of states \(M=10\).
  The dashed lines indicate the initial SNR bound, \eqref{eq:initbnd}, and the area bound, \eqref{eq:areabnd}, from \cite{Lahiri2013synapse}.}\label{fig:envproven}
\end{figure}

@@

Note that models with the serial topology must satisfy detailed balance.
Therefore, the conjecture from \cref{sec:finite} implies that the square-root bound of \cref{sec:sqrt} applies, implying the stricter envelope of \cref{eq:envDB}.
In \cref{fig:envnum} we compare these theoretical envelopes with the numerical maximisation from \cref{fig:envnum}.
We can see that this conjectured envelope is close to tight, being off by a constant factor.

\begin{figure}[tb]
  \centering
  % Requires \usepackage{graphicx}
  \includegraphics[width=0.8\linewidth]{LenvNum.pdf}\\
  \caption[Numerical and theoretical envelopes for the signal-to-noise ratio]
  {Numerical  and theoretical envelopes  for the signal-to-noise ratio, \eqref{eq:snrb}. 
  As in \cref{fig:envnum}, \cref{eq:finitelagrangian,eq:KTfinite,eq:finitederivs} are solved numerically with \(f\potdep=\frac{1}{2}\), the number of synapses \(N=1\) and the number of states \(M=10\).
  The orange line indicates maximisation of the memory curve over the space of all matrices satisfying the constraints \eqref{eq:constr}.
  The green line indicates maximisation over models with the serial topology, \ie potentiation only moves up one state and depression only moves down one state.
  The dotted blue line indicates the proven envelope, \cref{eq:env} and the solid blue line is the conjectured envelope assuming detailed balance, see \cref{eq:envDB}.}\label{fig:envnumth}
\end{figure}
